openTSNE: a modular \proglang{Python} library for t-SNE dimensionality reduction and embedding

### 2022-05-09 paper/main.tex
Update diagram
revision of implementation, mostly grammar

\newcommand{\opentsne}{\pkg{openTSNE}}

@@

%% - \author{} with primary affiliation
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{
  Pavlin G. Poli\v{c}ar\\
  Faculty of Computer and Information Science\\
  University of Ljubljana\\
  Ljubljana, Slovenia
  \AND
  Martin Stra\v{z}ar\\
  Broad Institute of MIT and Harvard\\
  Cambridge, MA, USA
  \AND
  Bla\v{z} Zupan\\
  Faculty of Computer and Information Science\\
  University of Ljubljana\\
  Ljubljana, Slovenia
}
\Plainauthor{P. G. Poli\v{c}ar, M. Stra\v{z}ar, B. Zupan}

@@

%% - \Abstract{} almost as usual
\Abstract{
One of the most popular techniques for constructing visualizations of large, high-dimensional data sets is $t$-distributed stochastic neighbor embedding (t-SNE). Recently, several extensions have been proposed to address scalability issues and the quality of the resulting visualizations. We introduce \opentsne, a modular \proglang{Python} library that implements the core t-SNE algorithm and its many extensions. The library is fast and can treat data sets containing millions of data points. \opentsne\ is open-source and is distributed under the BSD-3-Clause License. Its source code is available at \url{https://github.com/pavlin-policar/openTSNE}.
}

@@

The ever-growing volumes of high-dimensional data sets in machine learning call for efficient dimensionality reduction techniques and their implementations, capable of producing informative data visualizations. Popular approaches include principal component analysis, multidimensional scaling, $t$-distributed stochastic neighbor embedding (t-SNE)~\citep{maaten2008visualizing}, and uniform manifold approximation and projections (UMAP)~\citep{2018arXivUMAP}. Among these, t-SNE has received much attention as it can address high volumes of data and reveal the underlying data structure. For instance, t-SNE is widely used in the bioinformatics community in areas such as single-cell transcriptomics~\citep{macosko2015highly,cao2019single,tasic2018shared}, human genetics~\citep{hirata2019genetic}, metagenomic assembly~\citep{beaulaurier2018metagenomic}, the spatial organization of microbial communities~\citep{sheth2019spatial}, and metabolomics~\citep{tkachev2019differences}. To visually explore the data structure, the objects of interest, such as diseased and healthy tissues or single cells, are profiled through thousands of features that include, for example, expression of genes or concentration of metabolites. The task of t-SNE is to embed the objects of interest within a low, preferably two-dimensional embedding. Reports on single-cell gene expression data, our running example, often start with an overview of the cell landscape and use t-SNE to embed high-dimensional expression profiles into a two-dimensional space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b show two such embeddings.

\begin{figure*}[ht]
  \includegraphics[width=\textwidth]{macosko2015}
  \caption{\label{fig:macosko}
We use \opentsne\ to generate three t-SNE embeddings and demonstrate recent theoretical advances. The data in \textbf{(a)} and \textbf{(b)} includes 44,808 mouse retinal cells (dots in the plots) that are described with high-dimensional gene-expression profiles from \citet{macosko2015highly}. The data in \textbf{(c)} additionally contains 27,499 expression profiles from mouse retinal cells from \citet{shekhar2016comprehensive}. \textbf{(a)} We construct t-SNE embedding following the parameter choices from the original publication by \citet{maaten2008visualizing}. The visualization shows no preservation of the global organization of clusters, resulting from random initialization and an affinity model focused on preserving local neighborhoods. \textbf{(b)} A modern t-SNE embedding, utilizing the latest theoretical advances and practical recommendations constructed using a multi-scale affinity model, preserving both short-range and long-range interactions between data points and initialized so that the global layout is as meaningful as possible. Unlike in \textbf{(a)}, the green and blue clusters representing different sub-types of amacrine and bipolar cells are now localized to the same regions of the space, indicating a higher level of similarity than to other cell types. The embedding in \textbf{(c)} shows how existing t-SNE reference atlases can be used to place new samples into existing embeddings. The positions of new data points correspond to cell types from the reference atlas.
}
\end{figure*}

@@

However, easy access to efficient implementations almost universally correlates with the widespread adoption of novel computational techniques. Despite the many theoretical advancements, most popular t-SNE libraries have been slow to incorporate them into their implementations. In this paper, we present \opentsne, our open-source \proglang{Python} implementation of t-SNE, complete with its recently proposed extensions. \opentsne\ is easy to install and includes precompiled binaries available through popular \proglang{Python} package managers. It provides a familiar API, making it suitable as a drop-in replacement for existing t-SNE implementations.

@@

We introduce \opentsne, a comprehensive \proglang{Python} library that implements the t-SNE algorithm and its numerous recently proposed extensions. \opentsne\ is designed to accommodate all types of users and includes both a simple API that allows users to quickly obtain informative t-SNE visualizations, as well as a more flexible API that allows seasoned practitioners to modify all the
different aspects of the t-SNE algorithm.

The core t-SNE algorithm is comprised of three main steps, mirrored in the
structure of the \opentsne\ library:

\begin{enumerate}
\item Firstly, an affinity model describing the similarities between data points must be specified. Standard t-SNE predominantly deals with tabular data, and similarities are computed using Eqn.~\ref{eq:p_ij}. However, \citet{kobak2019art} showed that using mixtures of Gaussians instead of a single Gaussian often reveals better global consistency of the resulting visualization (albeit at some computational cost). Alternatively, one can speed up the overall computation by choosing a uniform kernel, which yields quantitatively similar results to standard t-SNE. \code{openTSNE.affinity} provides several affinity models to choose from but defaults to the standard perplexity-based model proposed in the original publication.

Furthermore, it is straightforward to define and use a custom affinity model. For instance, one can easily extend the \code{openTSNE.affinity.Affinities} class and provide an affinity model for relational data using a normalized adjacency matrix. This enables rapid experimentation and opens the doors to the visualization of non-tabular data.

\item Secondly, initial positions must be chosen for each data point in the embedding space. These can either be chosen randomly or can be used to impose some initial global structure on the embedding, e.g., using a PCA-based initialization. Accordingly, the \code{openTSNE.initialization} module provides several possible initialization schemes, or the user can also specify their own initialization by providing a \code{numpy.ndarray} object. By default, \opentsne\ opts for PCA-based initialization.

\item Finally, equipped with an affinity model and initial point positions, the only remaining step is to optimize the embedding such that the positions of the points in the embedding space reflect their corresponding similarities in the affinity model. In \opentsne, the initialization and affinities are used to construct an \code{openTSNE.TSNEEmbedding} object, which can then be optimized using different parameter settings. Once an embedding object is constructed, it can subsequently be optimized via the \code{.optimize} method.

Optimization can be performed using either the Barnes-Hut or FIt-SNE approximation scheme. While the FIt-SNE approximation scheme scales linearly with the number of samples, it introduces additional computational overhead, which is often unnecessary for smaller data sets. For smaller data sets, this sometimes results in longer runtimes than when using the Barnes-hut approximation scheme. By default, \opentsne\ uses a heuristic to ensure the fastest algorithm is chosen for the given data set, but this setting can also be specified by the user.
\end{enumerate}

@@

Embedding new data points to an existing \code{openTSNE.TSNEEmbedding}
follows the same three general steps as constructing a standard embedding:
\begin{enumerate}
\item Firstly, an affinity model must be provided describing the similarities between each new data point and the data points in the reference embedding. As before, \opentsne\ provides several possible affinity models in the \code{openTSNE.affinity} module. Most often, however, we want to utilize the same affinity model that was used to construct the reference embedding, as this leads to more spatially consistent embeddings. This is the default behavior of \opentsne.

\item Secondly, the initial positions of the new data points in the embedding space must be specified. These can, again, be initialized randomly or be informed by their similarity to the data points in the reference embedding. As before, \opentsne\ provides several initialization options in the \code{openTSNE.initialization} module, but the user is free to provide their own initialization.

\item Finally, what remains is to construct an embedding object, which can then be subject to optimization. Because this new embedding relies on an existing reference embedding, we construct a \code{openTSNE.PartialTSNEEmbedding} embedding object, which indicates its dependence on a reference \code{openTSNE.TSNEEmbedding} object. As before, \code{openTSNE.PartialTSNEEmbedding} inherits from \code{numpy.ndarray} and can therefore be easily manipulated. Optimization is, again, performed via the \code{.optimize} method.
\end{enumerate}

\begin{figure*}[t]
  \centering
  \includegraphics[width=0.75\textwidth]{opentsne-diagram}
  \caption{\label{fig:diagram}
  The high-level structure and flow of the \opentsne\ library.
  }
\end{figure*}

Fig.~\ref{fig:diagram} shows the high-level structure and flow of the \opentsne\ library.

Through different parameters, the user is given full control over every aspect of the t-SNE visualization algorithm, from different ways to define affinities to the delicate tuning of optimization parameters. This allows for rapid experimentation and enables users to quickly and efficiently inspect their data from numerous angles and at varying levels of resolution.

While \opentsne\ allows for an unprecedented level of control at every step, most end-users often only need a quick and easy way to create informative visualizations of their data sets. \opentsne\ provides a high-level API that closely follows the style of \pkg{scikit-learn}~\citep{sklearn_api}, a popular and widely used toolkit in the machine learning community. This allows users of all levels to quickly pick up and make use of the \opentsne\ library.

@@

\opentsne\ is designed to be easy to use and can be installed through either \pkg{PyPI} using \pkg{pip}
\begin{CodeChunk}
\begin{CodeInput}
pip install opentsne
\end{CodeInput}
\end{CodeChunk}
or using \pkg{conda} from the \pkg{conda-forge} channel
\begin{CodeChunk}
\begin{CodeInput}
conda install -c conda-forge opentsne
\end{CodeInput}
\end{CodeChunk}
\opentsne\ provides precompiled binaries for all major platforms (Windows, MacOS, Linux) for all currently supported \proglang{Python} versions (3.7---3.10). This is especially convenient for Windows users, as this obviates the need for a \proglang{C}/\proglang{C++} compiler, which is typically not bundled with the operating system.

To verify that that \opentsne\ was installed correctly, try to import the module:
\begin{CodeChunk}
\begin{CodeInput}
>>> import openTSNE
>>> print(openTSNE.__version__)
0.6.2
\end{CodeInput}
\end{CodeChunk}

@@

We demonstrate the simplest usage of \opentsne\ on the Iris data set from \citet{anderson1936species}. This small data set describes 150 iris flowers containing four variables (sepal length, sepal width, petal length, petal width) and their species (Iris Setosa, Iris Versicolor, Iris Virginica), which is the target variable. This data set is included in \pkg{scikit-learn}:
\begin{CodeChunk}
\begin{CodeInput}
>>> from sklearn import datasets
>>>
>>> iris = datasets.load_iris()
>>> iris.data[:3]
array([[5.1, 3.5, 1.4, 0.2],
       [4.9, 3. , 1.4, 0.2],
       [4.7, 3.2, 1.3, 0.2]])
\end{CodeInput}
\end{CodeChunk}
\pkg{scikit-learn} loads the data set into a \pkg{numpy} array, which can directly
be used by \opentsne:
\begin{CodeChunk}
\begin{CodeInput}
>>> import openTSNE
>>>
>>> embedding = openTSNE.TSNE().fit(iris.data)
>>> embedding[:3]
TSNEEmbedding([[-0.55581917, 17.56162344],
               [-1.76742536, 20.07763453],
               [-0.58444608, 20.00052495]])
\end{CodeInput}
\end{CodeChunk}
Fig.~\ref{fig:iris} shows a visualization of the resulting t-SNE visualization. As expected, t-SNE is correctly able to group together samples from the same species. Additionally, the t-SNE embedding reveals that Iris Setosa is markedly different from the other two species, while flowers from Iris Versicolor and Iris Virginica may appear more similar to one another.
\begin{figure*}[htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{iris}
  \caption{\label{fig:iris}
  A visualization of the obtained t-SNE embedding on the Iris data set. The colors
  correspond to different species of the Iris flower.
}
\end{figure*}
We can now use this embedding as a reference for new data points. In this simple example, we will duplicate some of the original samples and add them into the embedding.
\begin{CodeChunk}
\begin{CodeInput}
>>> new_data = iris.data[::3]
>>> new_embedding = embedding.transform(new_data)
>>> new_embedding[:3]
PartialTSNEEmbedding([[-0.02637324, 15.69956674],
                      [-2.84200792, 16.15478451],
                      [-2.4594136 , 15.37910876]])
\end{CodeInput}
\end{CodeChunk}
Fig.~\ref{fig:iris-transform} shows the resulting positions of the new data points, overlaid the original, reference embedding. As expected, t-SNE is correctly able to place the new data points to their corresponding groups.
\begin{figure*}[htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{iris-transform}
  \caption{\label{fig:iris-transform}
  A visualization of the original, reference data set (transparent points) overlaid with the new data points (opaque points).
}
\end{figure*}

@@

While the above example offers users a quick and easy way to generate informative visualizations, \opentsne\ allows us to control the t-SNE algorithm at every step. To demonstrate the flexibility of \opentsne, we will repeat the same analysis of the Iris data set as before, this time leveraging the more advanced features of the library. More practical examples of advanced usage will be shown in Section~\ref{sec:discussion}.

We, again, begin by loading the Iris data set using \pkg{scikit-learn}:
\begin{CodeChunk}
\begin{CodeInput}
>>> from sklearn import datasets
>>>
>>> iris = datasets.load_iris()
>>> iris.data[:3]
array([[5.1, 3.5, 1.4, 0.2],
       [4.9, 3. , 1.4, 0.2],
       [4.7, 3.2, 1.3, 0.2]])
\end{CodeInput}
\end{CodeChunk}
Next, we import the \opentsne\ library:
\begin{CodeChunk}
\begin{CodeInput}
>>> import openTSNE
\end{CodeInput}
\end{CodeChunk}
We now mirror the three steps listed above. First, we select an affinity model which describes the similarities between data points:
\begin{CodeChunk}
\begin{CodeInput}
>>> affinities = openTSNE.affinity.PerplexityBasedNN(iris.data, perplexity=30)
\end{CodeInput}
\end{CodeChunk}
Next, we determine the initial point positions in the embedding:
\begin{CodeChunk}
\begin{CodeInput}
>>> initialization = openTSNE.initialization.pca(iris.data)
>>> initialization[:3]
array([[-1.30971087e-04,  1.55848907e-05],
       [-1.32435711e-04, -8.63672049e-06],
       [-1.40967409e-04, -7.07276279e-06]])
\end{CodeInput}
\end{CodeChunk}
Next, using the affinity model and initialization, we can create an embedding object:
\begin{CodeChunk}
\begin{CodeInput}
>>> embedding = openTSNE.TSNEEmbedding(initialization, affinities)
>>> embedding[:3]
TSNEEmbedding([[-1.30971087e-04,  1.55848907e-05],
               [-1.32435711e-04, -8.63672049e-06],
               [-1.40967409e-04, -7.07276279e-06]])
\end{CodeInput}
\end{CodeChunk}
Notice that the embedding object contains the same point coordinates as the initialization. This is because although we have created an \code{openTSNE.TSNEEmbedding} object, we have not yet run any optimization. To optimize the embedding, we merely have to call the \code{embedding.optimize} method using our desired parameter settings. Here, we use the default parameter settings, including a shorter early-exaggeration phase, followed by a longer phase with no exaggeration.
\begin{CodeChunk}
\begin{CodeInput}
>>> embedding.optimize(250, exaggeration=12, momentum=0.5, inplace=True)
>>> embedding.optimize(500, momentum=0.8, inplace=True)
>>> embedding[:3]
TSNEEmbedding([[-0.55581917, 17.56162344],
               [-1.76742536, 20.07763453],
               [-0.58444608, 20.00052495]])
\end{CodeInput}
\end{CodeChunk}
We can validate that method produces the exact same embedding as before:
\begin{CodeChunk}
\begin{CodeInput}
>>> np.linalg.norm(openTSNE.TSNE().fit(iris.data) - embedding)
0.0
\end{CodeInput}
\end{CodeChunk}
Next, we can add new samples into the embedding:
\begin{CodeChunk}
\begin{CodeInput}
>>> new_embedding = embedding.prepare_partial(iris.data[::3], perplexity=5)
>>> new_embedding.optimize(
...     exaggeration=1.5, n_iter=250, learning_rate=0.1, momentum=0.8,
...     max_grad_norm=0.25, inplace=True
... )
>>> new_embedding[::3]
PartialTSNEEmbedding([[-0.02637324, 15.69956674],
                      [-2.84200792, 16.15478451],
                      [-2.4594136 , 15.37910876]])
\end{CodeInput}
\end{CodeChunk}
And validate that this also produces the exact same embedding as before:
\begin{CodeChunk}
\begin{CodeInput}
>>> np.linalg.norm(new_embedding - embedding.transform(iris.data[::3]))
0.0
\end{CodeInput}
\end{CodeChunk}

@@

Dimensionality reduction techniques implicitly assume that high-dimensional data
lies on a lower-dimensional manifold, which can accurately be captured by a small number of dimensions. However, there is no evidence that every data set can
accurately be described using only two dimensions, and any such embedding will
inevitably lead to a loss of information. Thus, it is beneficial to examine
multiple embeddings, each of which provides a different perspective on topology
and other data characteristics.

>>>>

\usepackage{xspace}
\newcommand{\opentsne}{\pkg{openTSNE}\xspace}

@@

%% - \author{} with primary affiliation
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{
  Pavlin G. Poli\v{c}ar\\
  Faculty of Computer and Information Science, University of Ljubljana\\
  Ljubljana, Slovenia
  \AND
  Martin Stra\v{z}ar\\
  Broad Institute of MIT and Harvard\\
  Cambridge, MA, USA
  \AND
  Bla\v{z} Zupan\\
  Faculty of Computer and Information Science, University of Ljubljana\\
  Ljubljana, Slovenia
}
\Plainauthor{P. G. Poli\v{c}ar, M. Stra\v{z}ar, B. Zupan}

@@

%% - \Abstract{} almost as usual
\Abstract{
One of the most popular techniques for visualizing large, high-dimensional data sets is $t$-distributed stochastic neighbor embedding (t-SNE). Recently, several extensions have been proposed to address scalability issues and the quality of the resulting visualizations. We introduce \opentsne, a modular \proglang{Python} library that implements the core t-SNE algorithm and its many extensions. The library is fast and can treat data sets containing millions of data points. We release \opentsne in open source and distribute it under the BSD-3-Clause License. Its source code is available at \url{https://github.com/pavlin-policar/openTSNE}.
}

@@

The ever-growing volumes of high-dimensional data sets in machine learning call for efficient dimensionality reduction techniques and their implementations to produce informative data visualizations. Popular approaches include principal component analysis, multidimensional scaling, $t$-distributed stochastic neighbor embedding (t-SNE)~\citep{maaten2008visualizing}, and uniform manifold approximation and projections (UMAP)~\citep{2018arXivUMAP}. Among these, t-SNE has received much attention as it can address high volumes of data and reveal the underlying data structure. For instance, t-SNE is widely used in the bioinformatics community in areas such as single-cell transcriptomics~\citep{macosko2015highly,cao2019single,tasic2018shared}, human genetics~\citep{hirata2019genetic}, metagenomic assembly~\citep{beaulaurier2018metagenomic}, the spatial organization of microbial communities~\citep{sheth2019spatial}, and metabolomics~\citep{tkachev2019differences}. To visually explore the data structure, the objects of interest, such as diseased and healthy tissues or single cells, are profiled through thousands of features that include, for example, the expression of genes or the concentration of metabolites. The task of t-SNE is to embed the objects of interest within a low, preferably two-dimensional embedding. Reports on single-cell gene expression data, our running example, often start with an overview of the cell landscape and use t-SNE to embed high-dimensional expression profiles into a two-dimensional space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b show two such embeddings.

\begin{figure*}[ht]
  \includegraphics[width=\textwidth]{macosko2015}
  \caption{\label{fig:macosko}
We use \opentsne to generate three t-SNE embeddings and demonstrate recent theoretical advances. The data in \textbf{(a)} and \textbf{(b)} includes 44,808 mouse retinal cells (dots in the plots) that are described with high-dimensional gene-expression profiles from \citet{macosko2015highly}. The data in \textbf{(c)} additionally contains 27,499 expression profiles from mouse retinal cells from \citet{shekhar2016comprehensive}. \textbf{(a)} We construct t-SNE embedding following the parameter choices from the original publication by \citet{maaten2008visualizing}. The visualization shows no preservation of the global organization of clusters, resulting from random initialization and an affinity model focused on preserving local neighborhoods. \textbf{(b)} A modern t-SNE embedding, utilizing the latest theoretical advances and practical recommendations constructed using a multi-scale affinity model, preserving both short-range and long-range interactions between data points and initialized so that the global layout is as meaningful as possible. Unlike in \textbf{(a)}, the green and blue clusters representing different sub-types of amacrine and bipolar cells are now localized to the same regions of the space, indicating a higher level of similarity than to other cell types. The embedding in \textbf{(c)} shows how existing t-SNE reference atlases can be used to place new samples into existing embeddings. The positions of new data points correspond to cell types from the reference atlas.
}
\end{figure*}

@@

However, easy access to efficient implementations almost universally correlates with the widespread adoption of novel computational techniques. Despite the many theoretical advancements, most popular t-SNE libraries have been slow to incorporate them into their implementations. In this paper, we present \opentsne, our open-source \proglang{Python} implementation of t-SNE, complete with its recently proposed extensions. \opentsne is easy to install and includes precompiled binaries available through popular \proglang{Python} package managers. It provides a familiar API, making it suitable as a drop-in replacement for existing t-SNE implementations.

@@

We introduce \opentsne, a comprehensive \proglang{Python} library that implements the t-SNE algorithm and its numerous recently proposed extensions. \opentsne includes a simple API that allows users to quickly obtain informative t-SNE visualizations and a more flexible API that allows seasoned practitioners to modify all the different aspects of the t-SNE algorithm.

The core t-SNE algorithm comprises three main steps, mirrored in the structure of the \opentsne library:

\begin{enumerate}
\item The t-SNE algorithm requires a specification of an affinity model to describe the similarities between data points. Standard t-SNE predominantly deals with tabular data and computes similarities using Eqn.~\ref{eq:p_ij}. However, \citet{kobak2019art} showed that using mixtures of Gaussians instead of a single Gaussian often reveals better global consistency of the resulting visualization (albeit at some computational cost). Alternatively, one can speed up the overall computation by choosing a uniform kernel, which yields quantitatively similar results to standard t-SNE. \code{openTSNE.affinity} provides several affinity models to choose from but defaults to the standard perplexity-based model proposed in the original publication.

In \opentsne, it is straightforward to define and use a custom affinity model. For instance, one can extend the \code{openTSNE.affinity.Affinities} class and provide an affinity model for relational data using a normalized adjacency matrix. Such extension enables rapid experimentation and opens the doors to the visualization of non-tabular data.

\item The optimization in t-SNE starts with initial positions for each data point in the embedding space. We can set the initial positions randomly or impose some initial global structure on the embedding with, for example, a PCA-based initialization. Accordingly, the \code{openTSNE.initialization} module provides several possible initialization schemes, The user can also specify their initialization by providing a \code{numpy.ndarray} object. By default, \opentsne opts for PCA-based initialization.

\item In optimizing the embedding, the t-SNE algorithm finds the positions of the points in the embedding space that reflect their corresponding similarities in the affinity model. \opentsne stores the initial positions and affinities in a \code{openTSNE.TSNEEmbedding} object that includes the \code{.optimize} method for embedding optimization that uses either the Barnes-Hut or FIt-SNE approximation scheme. While the FIt-SNE approximation scheme scales linearly with the number of samples, it introduces additional computational overhead, often unnecessary for smaller data sets, resulting in longer runtimes than the Barnes-hut approximation scheme. By default, \opentsne uses a heuristic to choose the fastest algorithm for a given data set.
\end{enumerate}

@@

Embedding new data points to an existing \code{openTSNE.TSNEEmbedding}
follows the same three general steps as constructing a standard embedding:
\begin{enumerate}
\item Firstly, we must provide an affinity model describing the similarities between each new data point and the data points in the reference embedding. As for the construction of the initial embedding, \opentsne provides several possible affinity models in the \code{openTSNE.affinity} module. Most often, however, we want to utilize the affinity model from the construction of the reference embedding, as this leads to more spatially consistent embeddings. Using an affinity model from reference embedding for the embedding of the new data is the default behavior of \opentsne.

\item Secondly, we must specify the initial positions of the new data points in the embedding space. We can, again, initialize these positions randomly or according to their similarity to the data points in the reference embedding. As before, \opentsne provides several initialization options in the \code{openTSNE.initialization} module, but the user is free to provide their specific initialization.

\item Finally, what remains is a construction of an embedding object, subject to optimization. Because this new embedding relies on an existing reference embedding, we construct a \code{openTSNE.PartialTSNEEmbedding} embedding object, which indicates its dependence on a reference \code{openTSNE.TSNEEmbedding} object. As before, \code{openTSNE.PartialTSNEEmbedding} inherits from \code{numpy.ndarray} and can therefore be easily manipulated. Optimization is, again, performed via the \code{.optimize} method.
\end{enumerate}

\begin{figure*}[htbp]
  \centering
  \includegraphics[width=0.9\textwidth]{opentsne-diagram}
  \caption{\label{fig:diagram}
  High-level structure and information flow of the \opentsne library. Solid black lines show the standard parameter flow, while red dashed lines indicate the utility of optional parameters, allowing more refined control over the t-SNE algorithm.
  }
\end{figure*}

Fig.~\ref{fig:diagram} shows the high-level structure and information flow of the \opentsne library. Through different parameters, the user is given complete control over every aspect of the t-SNE data visualization algorithm, from different ways to define affinities to the delicate tuning of optimization parameters. The parametrization of every part of t-SNE allows for rapid experimentation. It enables users to quickly and efficiently inspect their data from numerous angles and at varying levels of resolution.

While \opentsne allows for control at every step, most end-users often only need a quick and easy way to create informative visualizations of their data sets. \opentsne provides a high-level API that closely follows the style of \pkg{scikit-learn}~\citep{sklearn_api}, a popular and widely used machine learning toolkit.

@@

\opentsne is designed to be easy to use and can be installed through either Python Package Index \pkg{PyPI} using package installer \pkg{pip},
\begin{CodeChunk}
\begin{CodeInput}
pip install opentsne
\end{CodeInput}
\end{CodeChunk}
or using package manager \pkg{conda} from the \pkg{conda-forge} repository,
\begin{CodeChunk}
\begin{CodeInput}
conda install -c conda-forge opentsne
\end{CodeInput}
\end{CodeChunk}
\opentsne provides precompiled binaries for all major platforms (Windows, macOS, Linux) for all currently supported \proglang{Python} versions (3.7---3.10). Precompiled binaries are especially convenient for Windows users, as they obviate the need for a \proglang{C}/\proglang{C++} compiler, which is typically not bundled with Windows operating system.

To verify that that \opentsne was installed correctly, users can run Python and try to import the module by:
\begin{CodeChunk}
\begin{CodeInput}
>>> import openTSNE
>>> print(openTSNE.__version__)
0.6.2
\end{CodeInput}
\end{CodeChunk}

@@

We here outline the most straightforward usage of \opentsne on the Iris data set from \citet{anderson1936species}. This small data set profiles 150 iris flowers with four variables (sepal length, sepal width, petal length, petal width) and provides information about their species ({\em Iris Setosa}, {\em Iris Versicolor}, {\em Iris Virginica}). Iris data set comes with \pkg{scikit-learn}:
\begin{CodeChunk}
\begin{CodeInput}
>>> from sklearn import datasets
>>>
>>> iris = datasets.load_iris()
>>> iris.data[:3]
array([[5.1, 3.5, 1.4, 0.2],
       [4.9, 3. , 1.4, 0.2],
       [4.7, 3.2, 1.3, 0.2]])
\end{CodeInput}
\end{CodeChunk}
\pkg{scikit-learn} loads the data set into a \pkg{numpy} array which can directly be used by \opentsne:
\begin{CodeChunk}
\begin{CodeInput}
>>> import openTSNE
>>>
>>> embedding = openTSNE.TSNE().fit(iris.data)
>>> embedding[:3]
TSNEEmbedding([[-0.55581917, 17.56162344],
               [-1.76742536, 20.07763453],
               [-0.58444608, 20.00052495]])
\end{CodeInput}
\end{CodeChunk}
Fig.~\ref{fig:iris} displays the resulting t-SNE visualization. As expected, t-SNE can correctly group samples from the same species. The t-SNE embedding reveals that flowers from {\em Iris Setosa} are markedly different from the other two species. At the same time, flowers from {\em Iris Versicolor} and {\em Virginica} are more similar to one another.
\begin{figure*}[htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{iris}
  \caption{\label{fig:iris}
  A visualization of the constructed t-SNE embedding on the Iris data set. The colors correspond to different species of Iris.
}
\end{figure*}
We can now use the resulting embedding as a reference for new data points. In this simple example, we will duplicate some of the original samples and add them to the embedding.
\begin{CodeChunk}
\begin{CodeInput}
>>> new_data = iris.data[::3]
>>> new_embedding = embedding.transform(new_data)
>>> new_embedding[:3]
PartialTSNEEmbedding([[-0.02637324, 15.69956674],
                      [-2.84200792, 16.15478451],
                      [-2.4594136 , 15.37910876]])
\end{CodeInput}
\end{CodeChunk}
Fig.~\ref{fig:iris-transform} shows the resulting positions of the new data points, overlaid the original, reference embedding. As expected, t-SNE is correctly able to place the new data points to their corresponding groups.
\begin{figure*}[htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{iris-transform}
  \caption{\label{fig:iris-transform}
  A visualization of the original, reference data set (transparent points) overlaid with the new data points (opaque points).
}
\end{figure*}

@@

While the above example offers users a quick and easy way to generate informative visualizations, \opentsne allows us to control every step of the t-SNE embedding. To demonstrate this flexibility, we will repeat the exact analysis of the Iris data set as before, this time leveraging the library's more advanced features. We list other practical examples with more complex data sets in Section~\ref{sec:discussion}.

We, again, begin by loading the Iris data set using \pkg{scikit-learn}:
\begin{CodeChunk}
\begin{CodeInput}
>>> from sklearn import datasets
>>>
>>> iris = datasets.load_iris()
>>> iris.data[:3]
array([[5.1, 3.5, 1.4, 0.2],
       [4.9, 3. , 1.4, 0.2],
       [4.7, 3.2, 1.3, 0.2]])
\end{CodeInput}
\end{CodeChunk}
Next, we import the \opentsne library:
\begin{CodeChunk}
\begin{CodeInput}
>>> import openTSNE
\end{CodeInput}
\end{CodeChunk}
We now mirror the three t-SNE steps from the previous section. First, we select an affinity model which describes the similarities between data points:
\begin{CodeChunk}
\begin{CodeInput}
>>> affinities = openTSNE.affinity.PerplexityBasedNN(iris.data, perplexity=30)
\end{CodeInput}
\end{CodeChunk}
Next, we determine the initial point positions in the embedding:
\begin{CodeChunk}
\begin{CodeInput}
>>> initialization = openTSNE.initialization.pca(iris.data)
>>> initialization[:3]
array([[-1.30971087e-04,  1.55848907e-05],
       [-1.32435711e-04, -8.63672049e-06],
       [-1.40967409e-04, -7.07276279e-06]])
\end{CodeInput}
\end{CodeChunk}
Finally, using the affinity model and initialization, we can create an embedding object:
\begin{CodeChunk}
\begin{CodeInput}
>>> embedding = openTSNE.TSNEEmbedding(initialization, affinities)
>>> embedding[:3]
TSNEEmbedding([[-1.30971087e-04,  1.55848907e-05],
               [-1.32435711e-04, -8.63672049e-06],
               [-1.40967409e-04, -7.07276279e-06]])
\end{CodeInput}
\end{CodeChunk}
Notice that the embedding object contains the same point coordinates as the initialization: although we have created an \code{openTSNE.TSNEEmbedding} object, we have not yet run any optimization. To optimize the embedding, we have to call the \code{embedding.optimize} method using our desired parameter settings. We use the default parameter settings here, including a shorter early-exaggeration phase, followed by a longer phase with no exaggeration.
\begin{CodeChunk}
\begin{CodeInput}
>>> embedding.optimize(250, exaggeration=12, momentum=0.5, inplace=True)
>>> embedding.optimize(500, momentum=0.8, inplace=True)
>>> embedding[:3]
TSNEEmbedding([[-0.55581917, 17.56162344],
               [-1.76742536, 20.07763453],
               [-0.58444608, 20.00052495]])
\end{CodeInput}
\end{CodeChunk}
We can validate that method produces the exact same embedding as before:
\begin{CodeChunk}
\begin{CodeInput}
>>> np.linalg.norm(openTSNE.TSNE().fit(iris.data) - embedding)
0.0
\end{CodeInput}
\end{CodeChunk}
Next, we can add new samples into the embedding,
\begin{CodeChunk}
\begin{CodeInput}
>>> new_embedding = embedding.prepare_partial(iris.data[::3], perplexity=5)
>>> new_embedding.optimize(
...     exaggeration=1.5, n_iter=250, learning_rate=0.1, momentum=0.8,
...     max_grad_norm=0.25, inplace=True
... )
>>> new_embedding[::3]
PartialTSNEEmbedding([[-0.02637324, 15.69956674],
                      [-2.84200792, 16.15478451],
                      [-2.4594136 , 15.37910876]])
\end{CodeInput}
\end{CodeChunk}
and validate that this also produces the exact same embedding as before:
\begin{CodeChunk}
\begin{CodeInput}
>>> np.linalg.norm(new_embedding - embedding.transform(iris.data[::3]))
0.0
\end{CodeInput}
\end{CodeChunk}

@@

Dimensionality reduction techniques implicitly assume that high-dimensional data lies on a lower-dimensional manifold, which can accurately be captured by a small number of dimensions. However, there is no evidence that every data set can accurately be described using only two dimensions, and any such embedding will inevitably lead to a loss of information. Thus, it is beneficial to examine multiple embeddings, each providing a different perspective on topology and other data characteristics.

\begin{figure*}[htbp]
  \center
  \includegraphics[width=0.75\textwidth]{tasic2018}
  \caption{\label{fig:tasic}
  We use \opentsne to create four different
  visualizations of the \citet{tasic2018shared} data,
  each providing a different perspective into the topology of the data.
  The data set contains 21,874 single-cells originating from the mouse
  neocortex. Cluster annotations and colors are taken from the original
  publication. Warm colors correspond to excitatory neurons, cool colors
  correspond to inhibitory neurons, and gray/brown colors correspond to
  non-neuronal cells. Standard t-SNE \textbf{(a)} emphasizes local
  structure while increasing perplexity \textbf{(b)} results in a more
  meaningful layout of the clusters. We can also combine the two
  perplexities by using a multi-scale kernel affinity model \textbf{(c)}
  and obtain a trade-off between global and local structures.
  Alternatively, we can inspect more fine-grained structures and reveal
  smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
\end{figure*}
